%lambdlambdaa any author declaration will be ignored  when using 'plid' option (for double blind review)
\documentclass[pldi,nocopyrightspace]{sigplanconf}

% From PLDI Template {-{

% \usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage{amsmath}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

% }-}

% Setup {-{

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stmaryrd}
\usepackage{galois}
\usepackage{float}
\usepackage{latexsym}
\floatstyle{boxed}
\restylefloat{figure}
\usepackage[font=small,skip=0pt]{caption}

\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}

% This makes newlines not introduce paragraphs.
% Paragraphs must be explicitly marked with \par
% WHAT WIZARDRY IS THIS
\endlinechar=9\relax

% }-}

% Macros {-{

\newcommand\litM{M}
\newcommand\litTime{Time}
\newcommand\litStore{Store}
\newcommand\litVal{Val}
\newcommand\litKAddr{KAddr}
\newcommand\litKStore{KStore}
\newcommand\litEnv{Env}
\newcommand\litClo{Clo}

\newcommand{\itop}[1]{\operatorname{\mathit{#1}}}
\newcommand{\ttbfop}[1]{\operatorname{\mathtt{\boldsymbol{\mathbf{#1}}}}}
\newcommand{\ttbfbin}[1]{\mathbin{\mathtt{\mathbf{\boldsymbol{#1}}}}}
\newcommand{\ttop}[1]{\operatorname{\mathtt{#1}}}
\newcommand{\Concrete}[1]{\mathtt{\mathbf{#1}}}
\newcommand{\Abstract}[1]{\widehat{\mathtt{\mathbf{#1}}}}

% }-}

\begin{document}

% These are supposed to make align* not have breaks around it...
\setlength{\abovedisplayskip}{0em}
\setlength{\belowdisplayskip}{0em}
\setlength{\abovedisplayshortskip}{0em}
\setlength{\belowdisplayshortskip}{0em}

%\title{The Marriage of Monad Transformers and Abstract Interpreters: Modular Metatheory for Program Analysis}
\title{Galois Transformers and Modular Abstract Interpreters}
\subtitle{Reusable Metatheory for Program Analysis}
\authorinfo{David Darais}{Harvard University, USA}{darais@seas.harvard.edu}
\authorinfo{Matthew Might}{University of Utah, USA}{might@cs.utah.edu}
\authorinfo{David Van Horn}{University of Maryland, USA}{dvanhorn@cs.umd.edu}
\maketitle

% Abstract {-{
\begin{abstract}

The design and implementation of static analyzers have becoming increasingly
systematic.  In fact, for large classes of analyzers, design and implementation
have remained seemingly (and now stubbornly) on the verge of full mechanization
for several years.  A stumbling block in full mechanization has been the
\emph{ad hoc} nature of soundness proofs accompanying each analyzer. While
design and implementation is largely systematic, soundness proofs can change
significantly with what seem like minor changes to the semantics and analyzers
themselves.  An achievement of this work is to systematize, parameterize and
modularize the proofs of soundness, so as to make them composable across
analytic properties.

\par 

We solve the problem of systematically constructing static analyzers by
introducing the notion of a \emph{Galois transformer}, a monad transformer that
transports Galois connection properties.  In concert with a monadic
interpreter, we are able to define a library of monad transformers that
implement and represent classic context-, path-, and heap-(in-)sensitive
building blocks.  Moreover, these can be composed together to realize a program
analysis monad, \emph{independent of the language being analyzed}.

\par 

Significantly, a Galois transformer can be proved sound once and for all,
making it a reusable analysis component.  As new analysis features and
abstractions are developed and mixed in, soundness proofs need not be
reconstructed, as the composition of a monad transformer stack is sound by
virtue of its constituents.  Galois transformers provide a viable foundation
for reusable and composable metatheory for program analysis.

\par 

Finally, these Galois transformers shift the level of abstraction in analysis
design and implementation to a level where non-specialists have the ability to
synthesize sound analyzers over a number of parameters.

\end{abstract}
% }-}

\input{tmp/autogen/pldi15.markdown.tex}

\bibliographystyle{abbrvnat}
\bibliography{dvanhorn,davdar}

\end{document}
